### 2016-01-08 papers/CAFe/HIPS-paper/intro.tex

>>>>

A goal of high-level programming languages should be to \emph{allow} a programmer to
develop software for a specific hardware architecture, without actually \emph{exposing} hardware
features.  In fact, allowing low-level hardware details to be controlled by the programmer is actually
a barrier to software performance portability: the explicit use of low-level hardware details 
leads to code becoming excessively specialized to a given target system.  The path forward for high-level
programming languages should be to allow a programmer to provide more semantic information
about their \emph{intent}, thus allowing the compiler more freedom to choose how to instantiate this
intent in a specific implementation in order to retarget applications to new architectures.

Performance portability (or, even simply portability) has been a perpetual problem in computing,
especially within the high performance computing community in which relatively new and immature
emerging technologies are constantly being tried and tested.  As processor architectures hit limitations due
to physical and manufacturing constraints, we are seeing a trend towards more diversity in
processor designs than were experienced for the bulk of the 1990s and 2000s.  Portability, especially
performance portability, is becoming more challenging as moving to a new CPU may entail more than simple
tuning if the architecture is fundamentally different than previous ones.

A number of options are available to an HPC programmer beyond just using a serial
language with MPI or OpenMP.  New parallel languages have been under development for over a decade, 
for example, Chapel and the PGAS languages including Coarray Fortran (CAF).  Options for 
heterogeneous computing
include usage of OpenACC, CUDA, CUDA Fortran, and OpenCL for programming attached
accelerator devices.  Each of these choices provides the programmer with abstractions over parallel
systems that expose different levels of detail about the specific systems to compile to, and as such,
entail different degrees of effort to design for and realize performance portability.

% OpenACC directives and features:
%\begin{verbatim}
%● acc_copy(in|out)( ptr, bytes )
%● acc_create( ptr, bytes )
%● acc_delete( ptr, bytes )
%● acc_is_present( ptr, bytes )
%● acc_update_(device|local)( ptr, bytes )
%● acc_deviceptr( ptr )
%● acc_hostptr( devptr )
%● acc_[un]map_data( devptr, hostptr, bytes )
%● acc_memcpy_(to|from)_device
%\end{verbatim}

With the addition of coarrays in the 2008 standard, Fortran is now a \emph{parallel}
language.  Originally called Coarray Fortran (CAF) by its
authors\cite{Numrich:1998:CFP:289918.289920}, parallism in Fortran is similar to MPI in
that all cooperating CAF processes (called \emph{images}) execute a Single Program but with
different (Multiple) Data (SPMD).  Key concepts in CAF are parallel execution, distributed
memory allocation, and remote memory transfer.  Parallel execution and memory allocation
are straightforward as each program is replicated $NP$ times with each program image
allocating its own memory locally.  Special syntax is provided to indicate which arrays
are visible between distributed memory processes and remote memory transfer is indicated
with new array syntax using square brackets, for example,
\begin{verbatim}
   U(1) = U(1)[3]
\end{verbatim}
copies memory from the first element of \texttt{U} residing on image number 3 into
the corresponding element of the \emph{local} array \texttt{U}.  The fundamental programming
model is similar to that provided by message passing systems.

Since each CAF process executes the same program, there are no good avenues for exploiting
task-based parallelism.  Neither is there a mechanism for dynamic coarray memory
allocation, nor a mechanism for the remote execution of functions.  In CAF, like MPI, the
programmer often combines distributed memory processes running the same program image with a
shared memory threaded programming model using OpenMP (or OpenACC).  This model breaks down
when dealing with architectures that are not isomorphic to the model for which MPI and
relatives were designed -- a distributed collection of homogenous processing elements in
which each processing element supports a small number of parallel threads of execution.  For
example, an HPC system that makes use of traditional multicore CPUs as well as GPGPU accelerators
breaks this model and, to effectively program it, currently requires a mixture of programming
systems in a single application.

In this paper we propose CAFe: Co-Array Fortran Extensions.  These extensions to CAF introduce
relatively minor additions to the existing coarray portion of the Fortran language.  They
include the concept of subimages in order to nest coarray images, semantics for distributed
array declarations, memory placement, remote memory transfer, and remote task execution.  CAFe
builds upon the existing standardized coarray features of Fortran.  We will introduce the
CAFe extensions that are proposed to define their semantics, demonstrate their use in simple
examples, and provide experimental results demonstrating their performance in a prototype implementation.
CAFe is complementary to
previous work extending coarray Fortran\cite{mellor-crummey:2009:caf2,jin:2011:caf2}.

%Topics highlighted in this section are: 1. Distributed memory array allocation;
%2. Explicit memory placement; 3. Remote memory transfer; and 4. Remote execution.  This
%description is within the context of extensions to Fortran; as shorthand, these extensions
%are referred to as CAFe, for Coarray Fortran extensions.

%%% MJS: stopped editing here 1/7/2016
